---
title: "HMMexploration"
output:
  pdf_document: default
  html_document: default
date: "2023-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hidden Markov Model exploration for calves

# loading and visualizing data
load the libraries and the data from Matlab (it will already be chunked and averaged over a specified window)
```{r}
library(fitdistrplus)
library(momentuHMM)
library(data.table)
library(ggplot2)

all_other = read.csv('all_other.csv')
all_rest = read.csv('all_rest.csv')
all_travel = read.csv('all_travel.csv')

# rename the columns 
col_names = c('date_no', 'avg_speed', 'mean_resultant_length', 
              'max_depth', 'msa')
names(all_other) = col_names
names(all_rest) = col_names
names(all_travel) = col_names

all_all = rbind(all_rest, all_travel, all_other)
```

Look at the distributions of all 4 data streams

```{r, echo=FALSE}
par(mfrow=c(2,2))
hist(all_all$avg_speed, breaks=30)
hist(all_all$mean_resultant_length, breaks=30)
hist(all_all$max_depth, breaks=30)
hist(all_all$msa, breaks=30)
```

## parameter estimation for each variable

SPEED
```{r}
moment_match_gamma <- function(mu, sigma){
  # sigma is the standard deviation
  alpha = (mu ^ 2) / (sigma ^ 2)
  beta = mu / (sigma ^ 2)
  parameters = c(alpha, beta)
  names(parameters) = c('alpha', 'beta')
  parameters
}


# calculate the parameters for the travel state
travel_params = moment_match_gamma(
  mean(all_travel$avg_speed), 
  sd(all_travel$avg_speed)
)

#and for the rest state
rest_params = moment_match_gamma(
  mean(all_rest$avg_speed), 
  sd(all_rest$avg_speed)
)
```

plot the traveling and resting data and distributions

```{r, echo=FALSE}
# subplots
par(mfrow = c(2, 1))

# histogram for travel 
hist(all_travel$avg_speed, probability = TRUE, breaks=20, main = 'Travel', 
     xlim=c(0, max(all_travel$avg_speed)), xlab='')

# vertical line for mean
abline(v=mean(all_travel$avg_speed), col='red', lwd=2)

# curve for estimated gamma
curve(dgamma(x, travel_params['alpha'], travel_params['beta']), 
      add = T, col = 'blue', lwd = 2)

hist(all_rest$avg_speed, probability = TRUE, breaks=20, main = 'Rest',
     xlim=c(0, max(all_travel$avg_speed)))

abline(v=mean(all_rest$avg_speed), col='red', lwd=2)
curve(dgamma(x, rest_params['alpha'], rest_params['beta']), 
      add = T, col = 'blue', lwd = 2)

```
MEAN RESULTANT LENGTH
```{r}
moment_match_beta <- function(mu, var){
  # var is the variance
  alpha = ((1 - mu) / var - 1 / mu) * mu ^ 2
  beta = alpha * (1 / mu - 1)
  parameters = c(alpha, beta)
  names(parameters) = c('alpha', 'beta')
  parameters
}

travel_mrl_params = moment_match_beta(
  mean(all_travel$mean_resultant_length), 
  var(all_travel$mean_resultant_length)
)

rest_mrl_params = moment_match_beta(
  mean(all_rest$mean_resultant_length), 
  var(all_rest$mean_resultant_length)
)
```

plot the traveling and resting data and distributions

```{r}
par(mfrow = c(2, 1))
hist(all_travel$mean_resultant_length, probability = TRUE, breaks=20, main = 'Travel', 
     xlim=c(0, max(all_travel$mean_resultant_length)), xlab='')

# vertical line for mean
abline(v=mean(all_travel$mean_resultant_length), col='red', lwd=2)

# curve for estimated beta
curve(dbeta(x, travel_mrl_params['alpha'], travel_mrl_params['beta']), 
      add = T, col = 'blue', lwd = 2)

# ditto as above but for rest 
hist(all_rest$mean_resultant_length, probability = TRUE, breaks=20, main = 'Rest',
     xlim=c(0, max(all_travel$mean_resultant_length)))

abline(v=mean(all_rest$mean_resultant_length), col='red', lwd=2)

curve(dbeta(x, rest_mrl_params['alpha'], rest_mrl_params['beta']), 
      add = T, col = 'blue', lwd = 2)
```

DEPTH
```{r}
travel_depth_params = moment_match_gamma(
  mean(all_travel$max_depth), 
  sd(all_travel$max_depth)
)

rest_depth_params = moment_match_gamma(
  mean(all_rest$max_depth), 
  sd(all_rest$max_depth)
)
```

plot the traveling and resting data and distributions
```{r}
par(mfrow = c(2, 1))
hist(all_travel$max_depth, probability = TRUE, breaks=20, main = 'Travel', 
     xlim=c(0, max(all_travel$max_depth)), xlab='')

# vertical line for mean
abline(v=mean(all_travel$max_depth), col='red', lwd=2)

# curve for estimated gamma
curve(dgamma(x, travel_depth_params['alpha'], travel_depth_params['beta']), 
      add = T, col = 'blue', lwd = 2)

# ditto as above but for rest 
hist(all_rest$max_depth, probability = TRUE, breaks=20, main = 'Rest',
     xlim=c(0, max(all_travel$max_depth)))

abline(v=mean(all_rest$max_depth), col='red', lwd=2)

curve(dgamma(x, rest_depth_params['alpha'], rest_depth_params['beta']), 
      add = T, col = 'blue', lwd = 2)
```

AVERAGE MINIMUM SPECIFIC ACCELERATION
```{r}
travel_msa_params = moment_match_gamma(
  mean(all_travel$msa), 
  sd(all_travel$msa)
)

rest_msa_params = moment_match_gamma(
  mean(all_rest$msa), 
  sd(all_rest$msa)
)
```

plot the traveling and resting data and distributions
```{r}
par(mfrow = c(2, 1))
hist(all_travel$msa, probability = TRUE, breaks=20, main = 'Travel', 
     xlim=c(0, max(all_travel$msa)), xlab='')

# vertical line for mean
abline(v=mean(all_travel$msa), col='red', lwd=2)

# curve for estimated gamma
curve(dgamma(x, travel_msa_params['alpha'], travel_msa_params['beta']), 
      add = T, col = 'blue', lwd = 2)

# ditto as above but for rest 
hist(all_rest$msa, probability = TRUE, breaks=20, main = 'Rest',
     xlim=c(0, max(all_travel$msa)))

abline(v=mean(all_rest$msa), col='red', lwd=2)

curve(dgamma(x, rest_msa_params['alpha'], rest_msa_params['beta']), 
      add = T, col = 'blue', lwd = 2)
```

## simulate data and run simple HMMs

Simulate speed
```{r}
# simulate rest data from parameters
n_rest = nrow(all_rest)
sim_rest = rgamma(n_rest, rest_params['alpha'], rest_params['beta'])

# simulate travel data from parameters
n_travel = nrow(all_travel)
sim_travel = rgamma(n_travel, travel_params['alpha'], travel_params['beta'])

xlim = c(0, max(all_travel$avg_speed))

par(mfrow=c(2, 2))
hist(all_travel$avg_speed, breaks=30, xlim=xlim)
hist(all_rest$avg_speed, breaks=30, xlim=xlim)
hist(sim_travel, breaks=30, xlim=xlim)
hist(sim_rest, breaks=30, xlim=xlim)
```

compare simulated speed HMM output to real speed data HMM output
```{r}
#try the HMM on the simulated speed data
sim_speed<- c(sim_travel, sim_rest)
combined_sim_speeds <- data.frame(avg_speed = sim_speed)

sim_speed_data<- prepData(combined_sim_speeds, type='LL', coordNames=NULL)
dist<-list(avg_speed="gamma")
Par0 <- list(avg_speed= c(rest_params[['alpha']],travel_params[['alpha']] , rest_params[['beta']], travel_params[['beta']]))
stateNames <- c("resting","traveling")

sim_test1_speed <- fitHMM(sim_speed_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(sim_test1_speed)

#plot(sim_test1_speed, plotCI = TRUE, breaks=30)

```
The model did not work on the simulated data...

real speed data HMM
```{r}
all_speed<- c(all_travel$avg_speed, all_rest$avg_speed)
combined_avg_speeds <- data.frame(avg_speed = all_speed)

speed_data<- prepData(combined_avg_speeds, type='LL', coordNames=NULL)
dist<-list(avg_speed="gamma")
Par0 <- list(avg_speed= c(rest_params['alpha'],travel_params['alpha'] , rest_params['beta'], travel_params['beta']))
stateNames <- c("traveling","resting")

test1_speed <- fitHMM(speed_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(test1_speed)

states <- viterbi(test1_speed)
table(states)/nrow(speed_data) #well that didn't really work
plot(test1_speed, plotCI = TRUE, breaks=30)

```


Simulate MRL
```{r}
# simulate rest data from parameters
n_rest = nrow(all_rest)
sim_mrl_rest = rbeta(n_rest, rest_mrl_params['alpha'], rest_mrl_params['beta'])

# simulate travel data from parameters
n_travel = nrow(all_travel)
sim_mrl_travel = rbeta(n_travel, travel_mrl_params['alpha'], travel_mrl_params['beta'])

xlim = c(0, max(all_travel$mean_resultant_length))

par(mfrow=c(2, 2))
hist(all_travel$mean_resultant_length, breaks=30, xlim=xlim)
hist(all_rest$mean_resultant_length, breaks=30, xlim=xlim)
hist(sim_mrl_travel, breaks=30, xlim=xlim)
hist(sim_mrl_rest, breaks=30, xlim=xlim)
```

Compare simulated MRL results to real MRL data results
```{r}
sim_mrl<- c(sim_mrl_travel, sim_mrl_rest)
combined_sim_mrl <- data.frame(mrl = sim_mrl)

sim_mrl_data<- prepData(combined_sim_mrl, type='LL', coordNames=NULL)
dist<-list(mrl="beta")
Par0 <- list(mrl= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'] , rest_mrl_params['beta'], travel_mrl_params['beta']))
stateNames <- c("resting","traveling")

sim_test2_mrl <- fitHMM(sim_mrl_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(sim_test2_mrl)
plot(sim_test2_mrl, plotCI = TRUE, breaks=30)

```

real mrl data HMM
```{r}
all_mrl<- c(all_travel$mean_resultant_length, all_rest$mean_resultant_length)
combined_mrl <- data.frame(mrl = all_mrl)

mrl_data<- prepData(combined_mrl, type='LL', coordNames=NULL)
dist<-list(mrl="beta")
Par0 <- list(mrl= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'] , rest_mrl_params['beta'], travel_mrl_params['beta']))
stateNames <- c("resting","traveling")

test2_mrl <- fitHMM(mrl_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(test2_mrl)

states <- viterbi(test2_mrl)
table(states)/nrow(mrl_data) #well that didn't really work
plot(test2_mrl, plotCI = TRUE, breaks=30)
```

simulate max depth
```{r}
# simulate rest data from parameters
n_rest = nrow(all_rest)
sim_depth_rest = rgamma(n_rest, rest_depth_params['alpha'], rest_depth_params['beta'])

# simulate travel data from parameters
n_travel = nrow(all_travel)
sim_depth_travel = rgamma(n_travel, travel_depth_params['alpha'], travel_depth_params['beta'])

xlim = c(0, max(all_travel$max_depth))

par(mfrow=c(2, 2))
hist(all_travel$max_depth, breaks=30, xlim=xlim)
hist(all_rest$max_depth, breaks=30, xlim=xlim)
hist(sim_depth_travel, breaks=30, xlim=xlim)
hist(sim_depth_rest, breaks=30, xlim=xlim)
```

Compare simulated max depth HMM to real max depth HMM
```{r}
sim_depth<- c(sim_depth_travel, sim_depth_rest)
combined_sim_depth<- data.frame(max_depth = sim_depth)
sim_depth_data<- prepData(combined_sim_depth, type='LL', coordNames=NULL)
dist<-list(max_depth="gamma")
Par0 <- list(max_depth= c(rest_depth_params['alpha'],travel_depth_params['alpha'] , rest_depth_params['beta'], travel_depth_params['beta']))
stateNames <- c("traveling","resting")

sim_test2_depth <- fitHMM(sim_depth_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(sim_test2_depth)

states <- viterbi(sim_test2_depth)
plot(sim_test2_depth, plotCI = TRUE, breaks=30)
```
real max depth data HMM

```{r}
all_depth<- c(all_travel$max_depth, all_rest$max_depth)
combined_depth <- data.frame(max_depth = all_depth)
depth_data<- prepData(combined_depth, type='LL', coordNames=NULL)
dist<-list(max_depth="gamma")
Par0 <- list(max_depth= c(rest_depth_params['alpha'],travel_depth_params['alpha'] , rest_depth_params['beta'], travel_depth_params['beta']))
stateNames<- c("traveling","resting")

test2_depth <- fitHMM(depth_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(test2_depth)

states <- viterbi(test2_depth)
table(states)/nrow(depth_data) #well that didn't really work
plot(test2_depth, plotCI = TRUE, breaks=30)
```

simulate average MSA
```{r}
# simulate rest data from parameters
n_rest = nrow(all_rest)
sim_msa_rest = rgamma(n_rest, rest_msa_params['alpha'], rest_msa_params['beta'])

# simulate travel data from parameters
n_travel = nrow(all_travel)
sim_msa_travel = rgamma(n_travel, travel_msa_params['alpha'], travel_msa_params['beta'])

xlim = c(0, max(all_travel$msa))

par(mfrow=c(2, 2))
hist(all_travel$msa, breaks=30, xlim=xlim)
hist(all_rest$msa, breaks=30, xlim=xlim)
hist(sim_msa_travel, breaks=30, xlim=xlim)
hist(sim_msa_rest, breaks=30, xlim=xlim)
```

compare simulated HMMs to real HMMs

```{r}
sim_msa<- c(sim_msa_travel, sim_msa_rest)
combined_sim_msa<- data.frame(msa = sim_msa)
sim_msa_data<- prepData(combined_sim_msa, type='LL', coordNames=NULL)
dist<-list(msa="gamma")
Par0 <- list(msa= c(rest_msa_params['alpha'],travel_msa_params['alpha'] , rest_msa_params['beta'], travel_msa_params['beta']))
stateNames <- c("traveling","resting")

sim_test2_msa<- fitHMM(sim_msa_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(sim_test2_msa)
```

```{r}
all_msa<- c(all_travel$msa, all_rest$msa)
combined_msa <- data.frame(msa = all_msa)

msa_data<- prepData(combined_msa, type='LL', coordNames=NULL)
dist9<-list(msa="gamma")
Par09 <- list(msa= c(rest_msa_params['alpha'],travel_msa_params['alpha'] , rest_msa_params['beta'], travel_msa_params['beta']))
stateNames7 <- c("traveling","resting")

test2_msa <- fitHMM(msa_data, nbStates=2, dist=dist9, Par0=Par09, stateNames=stateNames7)
print(test2_msa)
```

The HMM on the average MSA data didn't work, basically the max log likelihood was infinity which is not good... maybe means that MSA is not a good summary statistic. 


## Multivariate HMMs

The best candidate variables seem to be the average speed, the mean resultant length, and the maximum depth for each 5 minute segment. 

A model with those 3 variables and 2 states: 
```{r}
var3_test<-all_all[,c(2,3,4)]

var3_test_data<-prepData(var3_test, type='LL',coordNames = NULL)

dist<-list(avg_speed="gamma",mean_resultant_length="beta",max_depth="gamma")

Par0 <- list( avg_speed=c(rest_params['alpha'],travel_params['alpha'] , rest_params['beta'], travel_params['beta']),
              mean_resultant_length= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'] , rest_mrl_params['beta'], travel_mrl_params['beta']), 
              max_depth=c(rest_depth_params['alpha'], travel_depth_params['alpha'], rest_depth_params['beta'], travel_depth_params['beta']))

stateNames <- c("traveling","resting")

test_3vars <- fitHMM(var3_test_data, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(test_3vars)
plot(test_3vars, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)
```
Now try a model with the same 3 variables and 3 states: resting, traveling, and other. The other state is when the video watchers were unsure if the animal was traveling or resting, so it is unclear yet biologically what this state is. 

```{r}
other_params = moment_match_gamma(
  mean(all_other$avg_speed), 
  sd(all_other$avg_speed)
)

other_mrl_params = moment_match_beta(
  mean(all_other$mean_resultant_length), 
  var(all_other$mean_resultant_length)
)

other_depth_params = moment_match_gamma(
  mean(all_other$max_depth), 
  sd(all_other$max_depth)
)

dist<-list(avg_speed="gamma",mean_resultant_length="beta",max_depth="gamma") #same as before
Par0 <- list( avg_speed=c(rest_params['alpha'],travel_params['alpha'] , other_params['alpha'], rest_params['beta'], travel_params['beta'], other_params['beta']),
              mean_resultant_length= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'], other_mrl_params['alpha'], rest_mrl_params['beta'], travel_mrl_params['beta'], other_mrl_params['beta']), 
              max_depth=c(rest_depth_params['alpha'], travel_depth_params['alpha'],other_depth_params['alpha'],  rest_depth_params['beta'], travel_depth_params['beta'], other_depth_params['beta']))

stateNames <- c("traveling", "other", "resting")

test_3vars_3st <- fitHMM(var3_test_data, nbStates=3, dist=dist, Par0=Par0, stateNames=stateNames)
print(test_3vars_3st)
plot(test_3vars_3st, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)
```

Now compare the AIC values of the models with different numbers of states
```{r}
AIC(test_3vars, test_3vars_3st, k = 2, n = NULL)
AICweights(test_3vars, test_3vars_3st, k = 2, n = NULL)
```
so the model with three states better fits the data, but this third state is maybe not that important

## Incorporating Covariates into the HMMS

Important covariates:
Individual identity (some whales might naturally behave differently from others)
  - this would affect the transition probabilities and the state distributions I think
Hour of day (there could be cycles in the whales behavior)
  - this would just affect the transition probabilities?
Age of the calf (as they age, calves might behave differently)
  - this would affect the transition probabilities and the state distributions I think

```{r}
calf_demo = read.csv('calf_demo_data.csv')

demo_hr<-calf_demo[,c(1,3,4,5,7)]

#try the model on this data set not broken up into traveling or resting
data_hr<-prepData(demo_hr, type='LL',coordNames = NULL, covNames=list(c("hour")))

dist<-list(avg_speed="gamma",mrl="beta",max_depth="gamma")

Par0 <- list( avg_speed=c(rest_params['alpha'],travel_params['alpha'] , rest_params['beta'], travel_params['beta']),
              mrl= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'] , rest_mrl_params['beta'], travel_mrl_params['beta']), 
              max_depth=c(rest_depth_params['alpha'], travel_depth_params['alpha'], rest_depth_params['beta'], travel_depth_params['beta']))

stateNames <- c("resting","traveling")

test_3vars_all <- fitHMM(data_hr, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames)
print(test_3vars_all)
plot(test_3vars_all, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)

states <- viterbi(test_3vars_all)
table(states)/nrow(data_hr)

states_df <- data.frame(State = states)
csv_file <- "calf_states.csv"
# Export the data frame to a CSV file
write.csv(states_df, file = csv_file, row.names = FALSE)
```


```{r}
formula <- ~ cosinor(hour, period = 24)

test_3vars_hr <- fitHMM(data_hr, nbStates=2, dist=dist, Par0=Par0, stateNames=stateNames, formula=formula)
print(test_3vars_hr)
plot(test_3vars_hr, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)

AIC(test_3vars_all, test_3vars_hr, k = 2, n = NULL)
AICweights(test_3vars_all, test_3vars_hr, k = 2, n = NULL)
```
According to the AIC values, the model with the cosinor 24 hour cycle as a covariate for the transition probabilities is better. 

Now trying a model with 3 states, 3 variables, and the hour as a covariate on the transition probabilities
```{r}

dist<-list(avg_speed="gamma",mrl="beta",max_depth="gamma") #same as before
Par0 <- list( avg_speed=c(rest_params['alpha'],travel_params['alpha'] , other_params['alpha'], rest_params['beta'], travel_params['beta'], other_params['beta']),
              mrl= c(rest_mrl_params['alpha'],travel_mrl_params['alpha'], other_mrl_params['alpha'], rest_mrl_params['beta'], travel_mrl_params['beta'], other_mrl_params['beta']), 
              max_depth=c(rest_depth_params['alpha'], travel_depth_params['alpha'],other_depth_params['alpha'],  rest_depth_params['beta'], travel_depth_params['beta'], other_depth_params['beta']))

stateNames <- c("resting", "traveling", "other")
formula <- ~ cosinor(hour, period = 24)

test_3vars_3st <- fitHMM(data_hr,nbStates=3, dist=dist, Par0=Par0, stateNames=stateNames, formula=formula)
print(test_3vars_3st)
plot(test_3vars_3st, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)

AIC(test_3vars_3st, test_3vars_hr, k = 2, n = NULL)
AICweights(test_3vars_3st, test_3vars_hr, k = 2, n = NULL)
#the model with 3 states is better based on AIC values

states <- viterbi(test_3vars_3st)
table(states)/nrow(data_hr)
states_df <- data.frame(State = states)
csv_file <- "calf_3states.csv"
# Export the data frame to a CSV file
write.csv(states_df, file = csv_file, row.names = FALSE)
```



Now I will try to incorporate age as a covariate on the transition probabilities, and as a covariate of the probability distributions for the average speed, mean resultant length, and maximum depth. I expect that the calves will travel faster, go deeper, and I'm not sure how it could affect the mrl but it might?

I could look at simple regressions to see if age affects the overall average of each variable
```{r}
library(dplyr)
drone_data = read.csv('tagged_animals_summary.csv')
drone_data <- drone_data %>%
  rename(ID = CATS_Tag_No)
calf_demo <- calf_demo %>%
  left_join(drone_data, by = "ID")

data_age<-calf_demo[,c(1,4,5,7,13)]
#want to get rid of the rows with NA values for the DSB
data_age <- na.omit(data_age)

#do some averaging of the speed, mrl, and max_depth, and do some linear regressions
avg_speed_summary <- data_age %>%
  group_by(ID) %>%
  summarise(Avg_Avg_Speed = mean(avg_speed))
merged_data <- inner_join(avg_speed_summary, drone_data, by = "ID")

ggplot(merged_data, aes(x = DSB, y = Avg_Avg_Speed)) +
  geom_point() +
  labs(title = "Scatter Plot of Average Speed vs. Age",
       x = "Age",
       y = "Average Speed")



```
The speed actually looks like it is decreasing with age?

```{r}
#do some averaging of the speed, mrl, and max_depth, and do some linear regressions
avg_mrl_summary <- data_age %>%
  group_by(ID) %>%
  summarise(Avg_mrl = mean(mrl))
merged_data2 <- inner_join(avg_mrl_summary, drone_data, by = "ID")

ggplot(merged_data2, aes(x = DSB, y = Avg_mrl)) +
  geom_point() +
  labs(title = "Scatter Plot of Average mean resultant length vs. Age",
       x = "Age",
       y = "Avg mean resultant length")
```
Looks pretty random

```{r}
#do some averaging of the speed, mrl, and max_depth, and do some linear regressions
avg_depth_summary <- data_age %>%
  group_by(ID) %>%
  summarise(Avg_depth = mean(max_depth))
merged_data3 <- inner_join(avg_depth_summary, drone_data, by = "ID")

ggplot(merged_data3, aes(x = DSB, y = Avg_depth)) +
  geom_point() +
  labs(title = "Scatter Plot of Average max depth vs. Age",
       x = "Age",
       y = "Avg max depth")
```
Looks kind of positive 


I got the covariates to work!!! And the AIC value is way lower  -2807.954 versus  -3417.372  for the model without the whale age as a covariate
```{r}
data_age_prep<-prepData(data_age, type='LL',coordNames = NULL, covNames=list(c("DSB")))
speedDM<-list(mean=~DSB, sd=~DSB)
depthDM<-list(mean=~DSB, sd=~DSB)
mrlDM<-list(shape1=~1, shape2=~1)
dist<-list(
  avg_speed="gamma",
  mrl="beta",
  max_depth="gamma"
  )

Par0 <- list( 
  avg_speed = c(
    rest_params['alpha'], travel_params['alpha'], 
    rest_params['beta'],  travel_params['beta']
  ),
  mrl= c(
    rest_mrl_params['alpha'], travel_mrl_params['alpha'], 
    rest_mrl_params['beta'],  travel_mrl_params['beta']
  ), 
  max_depth=c(
    rest_depth_params['alpha'], travel_depth_params['alpha'], 
    rest_depth_params['beta'],  travel_depth_params['beta']
  )
)

#have to do this so the Par and DM match or something
Par1<-getParDM(data = data_age_prep, 
               nbStates = 2, 
               dist = list(avg_speed="gamma",mrl="beta",max_depth="gamma"), 
               Par=Par0, 
               DM=list(avg_speed=list(mean=~DSB, sd=~DSB), mrl=list(shape1=~1, shape2=~1), max_depth=list(mean=~DSB, sd=~DSB))
               )

stateNames <- c("traveling","resting")

test_cov <- fitHMM(
  data_age_prep, 
  nbStates=2, 
  dist=dist, 
  DM=list(
    avg_speed=speedDM, 
    mrl=mrlDM, 
    max_depth=depthDM), 
  Par0=Par1, 
  stateNames=stateNames
)
print(test_cov)
plot(test_cov, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)

```




Now I want to try the best possible model, that has the age as a covariate on the transition probabilities and the state distributions, and the hour as a covariate on the transition probabilities.

```{r}
data_age_hr<-calf_demo[,c(1,3,4,5,7,13)]
#want to get rid of the rows with NA values for the DSB
data_age_hr <- na.omit(data_age_hr)

data_aghr_prep<-prepData(data_age_hr, type='LL',coordNames = NULL, covNames=c("DSB","hour"))

#Effects of the covariates on the variable probability distributions
speedDM<-list(mean=~DSB, sd=~DSB)
depthDM<-list(mean=~DSB, sd=~DSB)
mrlDM<-list(shape1=~1, shape2=~1)

#BOTH covariates effects on the transition probabilities
formula <- ~ DSB + cosinor(hour, period = 24)

dist<-list(
  avg_speed="gamma",
  mrl="beta",
  max_depth="gamma"
  )

Par0 <- list( 
  avg_speed = c(
    rest_params['alpha'], travel_params['alpha'], 
    rest_params['beta'],  travel_params['beta']
  ),
  mrl= c(
    rest_mrl_params['alpha'], travel_mrl_params['alpha'], 
    rest_mrl_params['beta'],  travel_mrl_params['beta']
  ), 
  max_depth=c(
    rest_depth_params['alpha'], travel_depth_params['alpha'], 
    rest_depth_params['beta'],  travel_depth_params['beta']
  )
)

#have to do this so the Par and DM match or something
Par1<-getParDM(data = data_aghr_prep, 
               nbStates = 2, 
               dist = list(avg_speed="gamma",mrl="beta",max_depth="gamma"), 
               Par=Par0, 
               DM=list(avg_speed=list(mean=~DSB, sd=~DSB), mrl=list(shape1=~1, shape2=~1), max_depth=list(mean=~DSB, sd=~DSB))
               )

stateNames <- c("traveling","resting")

test_2cov <- fitHMM(
  data_aghr_prep, 
  nbStates=2, 
  dist=dist, 
  DM=list(
    avg_speed=speedDM, 
    mrl=mrlDM, 
    max_depth=depthDM), 
  Par0=Par1, 
  stateNames=stateNames,
  formula=formula
)
print(test_2cov)
plot(test_2cov, plotCI = TRUE, breaks=30, sepStates=TRUE, ask=TRUE)


```












